---
title: "P8124 Assignment 1 Question 5"
author: "Christine Lucille Kuryla"
date: "2023-09-19"
output: html_document
---

## Problem 5: DAGs and the dagitty library

###Construct DAG
```{r construct_dag}
# Problem 5

library(dagitty)

# Construct DAG 

dag_q5 <- dagitty('dag {
    D [pos="0,1"]
    E [pos="1,1"]
    A [pos="2,1"]
    B [pos="3,1"]
    F [pos="4,1"]
    C [pos="1,0"]
    G [pos="4,0"]
    H [pos="5,0"]

    D -> E -> A <- B <- G -> H
    E -> F -> H
    E <- C -> H
    C -> B
    C -> F -> G
    E -> F

}')

plot(dag_q5)
```

### 5 a) List all paths from C to H:

```{r a}
# List all paths from C to H

paths(dag_q5, from = "C", to = "H")
```

### b) Determine whether E is d-separated from G given A,B
```{r b}
# b) Determine whether E is d-separated from G given A,B
dseparated(dag_q5, X = "E", Y = "G", Z = c("A", "B"))
paths(dag_q5, "E", "G", c("A", "B"))
```
E is NOT d-separated from G given {A,B}.


### c) Conditional independencies
```{r c}
# c) List the conditional independencies implied by the model with and without the option "type = all.pairs"
# What explains the difference between the two lists, why is the first one shorter than the second?

# Conditional independencies without the option type = "all.pairs"
impliedConditionalIndependencies(dag_q5)

# Conditional independencies with the option type = "all.pairs"
impliedConditionalIndependencies(dag_q5, type = "all.pairs")
```

The two lists are different because XXXXXXXXXXXXXXXX

### d) Simulation, Markov blanket verification, regression

```{r d}
# d) Simulate the DAG using simulateSEM(), which associates the DAG with a linear structural equation model. 
# Verify that the Markov blanket property holds for vertex B.
# Do this by examining the p-values (or CIs) for all covariates outside the Markov blanket in the regression of B ~ Mb(B,G) + remaining covariates

# Simulate DAG
sim_sem <- simulateSEM(dag_q5,
            b.lower = -0.7,
            b.upper = 0.7,
            N = 500)

# Verify Markov blanket property holds for vertex B

# Markov blanket for B
markovBlanket(dag_q5, v = "B")

# Regression of B ~ Mb(B,G) + remaining covariates
summary(glm(data = sim_sem, formula = B ~ C + G + A + E + D + F + H))
```

We see that the Markov blanket on the DAG G for vertex B is {C, G, A, E}. This is verified by the regression on the vertices such that the p-value for the variables representing the verticies in the Markov blanket (C, G, A, E) all have very small p-values with asterisks indicating as such, while the remaining covariates have larger, non-significant p-values, indicating that XXXXXXXXXXXXX.